---
title: "Homework 3 Suglia"
author: "Elena Suglia"
date: "11/17/2019"
output: pdf_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(rjags)
library(brms)
library(MCMCpack)
library(lattice)
library(dplyr)
library(MCMCglmm)
library(rstan)
library(lme4)
```

# Question 1
*Using the “litterbags.csv” data set, create and run a model in JAGS that corresponds to this one:*

```{r, eval = FALSE}
lmer(N_min_rate~Celastrus + (1|Plot), data=litterbags)
```

*This data set contains nitrogen mineralization rates at 7 different plots where the treatment was whether or not the invasive liana Celastrus orbiculatus was present or not.*

*In your answer, please include the model code and the means and standard deviations of the intercept, slope, group-level (i.e. among-plot) variance and individual-level (within-plot) variance. Briefly report what you did to check model convergence.*

## Model code in text file

```{r, eval = FALSE}
 model {
  
   # Likelihood
  for (i in 1:n) {
    y[i] ~ dnorm(y.hat[i], tau.y)
    y.hat[i] <- a[Plot[i]] + b*x[i] # regression with intercept for each plot
  }
  
  # group-level model
  for (j in 1:7) { # 7 plots
    a[j] ~ dnorm(mu.a, tau.a) # separate intercepts for every plot, no pooling of information
  }
  
  #priors
  mu.a ~ dnorm(0, 0.1) # overall mean
  sigma.a ~ dnorm(0, 0.5)T(0,) # weakly informative prior on the group-level standard deviation
  tau.a <- pow(sigma.a, -2) # group-level precision
  
  # individual-level priors
  b ~ dnorm(0, 0.1) #  prior for slope
  sigma.y ~ dnorm(0, 0.1)T(0,) # weakly informative prior on the individual-level sd
  tau.y <- pow(sigma.y, -2) # individual-level precision
}
```

## Read in data
```{r, message = FALSE, warning = FALSE}
d = read_csv("litterbags.csv")
```


## Define data needed for model

```{r, message = FALSE, warning = FALSE, results = 'hide'}
length(unique(d$Plot)) # 7 plots
d2.data <- list(n=nrow(d), y=d$N_min_rate, x=d$Celastrus,
plot.index=d$Plot, n.plots=length(unique(d$Plot)))
```

## Have jags generate starting values

```{r, message = FALSE, warning = FALSE, results = 'hide'}
d2.inits <- list(list(sigma.a = 1, sigma.y=2), list(sigma.a = 2,
sigma.y=1), list(sigma.a = 5, sigma.y=0.4))
d2 <- jags.model("litterbags_jags_hw3.txt", data=d2.data, inits = d2.inits,
n.chains=3, n.adapt=1000)
```

## Do a trial run with some parameters, and check convergence

```{r, message = FALSE, warning = FALSE}
d2.samp <- coda.samples(d2, c("b", "sigma.y", "mu.a", "sigma.a"), n.iter=1000)
summary(d2.samp)
par(mar=rep(2, 4))
plot(d2.samp)
acfplot(d2.samp)
gelman.diag(d2.samp)
```

## Add more iterations and include all parameters of interest

```{r, message = FALSE, warning = FALSE}
d2.samp <- coda.samples(d2, c("a", "b", "sigma.y", "mu.a", "sigma.a", "y.hat"), n.iter=5000, thin=5)
```

## Look at the results

```{r, message = FALSE, warning = FALSE}
d2.summ <- summary(d2.samp) 
d2.stats <- as.data.frame(d2.summ$statistics)
head(d2.stats, 100)
y.hat.rows <- grep("y.hat", rownames(d2.stats))
y.hat <- d2.stats$Mean[y.hat.rows] # To get fitted values of the model, look at the stats and pull out the rows corresponding to y.hat
resids <- d$N_min_rate - y.hat  
# for convenience, we can then put those values into our data frame and examine them
d <- cbind(d, y.hat = y.hat, resid = resids)
```

## Some summary statistics:
## The means and standard deviations of:

- slope = b
- intercept = mu.a
- group-level (i.e. among-plot) variance = sigma.a
- individual-level (within-plot) variance = sigma.y

```{r, echo = FALSE}
woo = summary(d2.samp)[1]
woo$statistics[,c('Mean', 'SD')]
```

## See how well the model ran

```{r, message = FALSE, warning = FALSE}
qqnorm(d$resid)
boxplot(resid~Plot, data=d)
# Now that random intercepts for county are included in the model, does it look like the residuals differ strongly by county? 
```

## Look at DIC

```{r, message = FALSE, warning = FALSE}
d2.DIC <- dic.samples(d2, n.iter=5000, thin=5)
d2.DIC
```

# For comparison, let's quickly fit the complete-pooling model that has no county effects. 
#d1.data <- list(n=nrow(d), y=d$N_min_rate, x=d$Celastrus)
# for this simple model we can let jags generate starting values
#d1 <- jags.model("d_completepool_jags.txt", data=d1.data, n.chains=3, n.adapt=1000)
#d1.samp <- coda.samples(d1, c("a", "b", "sigma.y"), n.iter=1000)
#d1.DIC <- dic.samples(d1, n.iter=1000)


# Question 2
*Fit a hierarchical (multilevel) model in JAGS using the data sets “immunity.csv” and “patient_age.csv”. The model should include “immune.level” as the response variable, “time” as an individual-level predictor, and “age” as a group-level predictor.*

*These data are repeated measures on individual patients, so each patient is a “group” in the data set. Each row in the data set “immunity.csv” is one observation on one patient. Each row in the data set “patient_age.csv” is the age of each patient at time the study began – so this is a group level predictor, with one row of data per patient. In other words, here “age” is analogous to county-level bedrock uranium content in the Gelman & Hill radon example.*

*Hints: You can use the column “patient” in “immunity.csv” to index the random intercept for patient. You can then use the column “age” in “patient_age.csv” in the group-level regression that explains some of the variation in the random intercepts. There is a model like this on page 361 of Gelman & Hill.*  

*In your answer, include the model you created. Also report the means and standard deviations of:*
- *the slopes of the individual-data-point-level and “group”-level (i.e. patient-level) regressions*
- *the individual-data-point-level and “group”-level variance parameters*
